---
title: "06_Phenotype_analysis.Rmd"
author: "R. Burke Squires"
date: "8/15/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Looks pretty good. Now what? flowType!

If you don't have flowType, i.e. if library(flowType) doesn't work):

(Ideally you can run this during lunch)

#source("http://bioconductor.org/biocLite.R")
#biocLite('rrcov')
#biocLite('codetools')
#biocLite('foreach')
#biocLite('flowMerge')
#biocLite('flowType')
#biocLite('RchyOptimyx')

```{r}
library(flowType)
library(RchyOptimyx)
```

```{r}
# For convinience, rename the channels of the flowSet to more phenotype-friendly names: 
colnames(viable.fs) <- c("FSC-A", "FSC-H", "SSC-A", "KI67", "CD3", "CD8", "CD4", "Dump", "CD127")
```

# Now we run flowType on a single flowFrame to see how it works. It is hard!
# Frame is your flowFrame object.
# PropMarkers are the indices or markers which you want to involve in the analysis. For us, CD4 is the 7th marker in the list above, CD8 is the 6th, and KI67 is the 4th. How you order them does not matter as long as you are consistent.
# Methods are the gate threshold values in the same order as above.
# MarkerNames is the full vector of all channel names.
```{r}
ft1 <- flowType(Frame = viable.fs[[1]], 
                PropMarkers=c(7, 6, 4), 
                Methods='kmeans', 
                MarkerNames=colnames(viable.fs))
```
# Examine ft1 -- see what's in there.
# In the console below type f1@ and then press your tab key to see available components to explore!


# Next, use fsApply to compute all phenotypes for thew hole flow set.
# Notice that we have our own function defined within the fsApply call -- you can do this if your function is so short, that you don't need to define it separately. Also notice the '/nrow(x)' part -- this is so that instead of cell counts we get cell proportions. We cannot use cell counts because the total number of starting cells is different for the different samples.
```{r}
ft <- fsApply(viable.fs, 
              function(x) flowType(x,
                                   PropMarkers=c(7, 6, 4), 
                                   Methods='kmeans', 
                                   MarkerNames=colnames(viable.fs))@CellFreqs/nrow(x))
rownames(ft) <- sampleNames(viable.fs)
```

# We want to identify phenotypes which separate our data into two groups based on survival times. We can calculate some p-values to use as a guage on the phenotypes' importance.
# First, let's remove samples with very low live CD3+ counts (say < 1000 cells):
```{r}
remove.low <- which(as.numeric(fsApply(viable.fs, nrow)) < 1000)
ft <- ft[-remove.low, ]
```

# Keep track of survival time before removing low viable count samples just in case we need it later
```{r}
full.survival.data <- survival
survival <- survival[-remove.low]
```

# Now identify the patients with survival less than 1000 days and over (from an earlier plot this looks like the dividing number!)
```{r}
group1 <- which(survival < 1000)
group2 <- which(survival > 1000)
```

# Calculate the p-values. Here is the p-value for a single phenotype:
(this will give us an error when run outside of a loop)
```{r}
one.pval <- t.test(ft[group1, "CD4-CD8+"], ft[group2, "CD4-CD8+"])$p.value
```

# Use a for loop to calculate the p-values by looping over all phenotypes.
# Initialize a vector of p-values to be all 1:
```{r}
pvals <- rep(1, ncol(ft)) # ncol is the number of phenotypes!
for (i in 1:ncol(ft)){
  if (sd(ft[, i]) == 0) {# if no variation in the phenotype measurement, the p-value will be undefined.
    pvals[i] <- 1
  } else {
    pvals[i] <- t.test(ft[group1, i], ft[group2, i])$p.value
  }
}
names(pvals) <- colnames(ft)
pvals
```

# Now we can use the p-values as a way to score each phenotype's importance. However, we want to find the phenotype with the fewest number of markers (most robust, efficient, cheap to make into a panel) without losing much of the ability to separate the patients with low survival time from those with high survival time. Instead of 'pvals' though, let's use -log10 (pvals). Typically the lower the p-value is, the better the phenotype is. By taking the -log10 of the p-value as the score, we can now say the higher the score -- the better the phenotype.

# First, RchyOptimyx wants us to provide all possible phenotype combinations in terms of "+" or "-" combinations of markers. Here we use '0' for a negative expression of a marker, '1' if the marker is neutral (not involved in the phenotype at all), and '2' for a positive expression of the marker. 'Signs' contains this required variable. (very computer science-y, but we are working on an easier version of flowType and RchyOptimyx right now!)
```{r}
library(sfsmisc)
Signs <- t(digitsBase(1:(3^3-1), 3, ndigits=3))
rownames(Signs) <- colnames(ft)
colnames(Signs) <- c("CD4", "CD8", "KI67")
```

# Now to run RchyOptimyx. Here are a couple of additional required parameters:
# startPhenotype: we don't necessarily want RchyOptimyx to check ALL Possible phenotypes. We can specify one with a low p-value that we want it to reach by combining its constituent markers and their espression.

# For example, "012" means CD4-CD8(neutral)KI67+ == CD4-KI67+.

# To get a broader view, we should not simply pick the one with the lowest value. We can generate a few RchyOptimyx trees and then merge them into one plot:
```{r}
rch1 <- RchyOptimyx(Signs, -log10(pvals), startPhenotype = "012", trimPaths=FALSE, pathCount=6)
rch2 <- RchyOptimyx(Signs, -log10(pvals), "002", trimPaths=FALSE, pathCount=6)
rch3 <- RchyOptimyx(Signs, -log10(pvals), "202", trimPaths=FALSE, pathCount=6)
merged <- merge(rch1, merge(rch2, rch3)) # Can only merge two at a time!
```

# Instead of plotting in RStudio, let's write this to a .pdf file!
# Specify file name:
```{r}
pdf('./Documents/Workshop/rchy.pdf')
# Do your plotting:
plot(merged, phenotypeScores=-log10(pvals))
# Tell R you are done plotting and it is safe to write the information to the file.
dev.off()
```

# Let's also save the flowType results and p-values:
```{r}
results <- rbind(ft, pvals)
rownames(results)[nrow(results)] <- "P-values (Uncorrected)"
write.csv(results, file="/home/rguru/Documents/Workshop/results.csv")
```

# Navigate to that folder to see the plot and .csv report!

# See that the 'best' phenotype is KI67+
```{r}
ft[, "KI67+"]*100
par(mfrow=c(1,1))
boxplot(ft[group1, "CD4-KI67+"]*100, ft[group2, "CD4-KI67+"]*100, boxwex=0.2, labels=c("Group1", "Group2"))
```
# NOT great at all, but to be expected with such a small set of samples and small set of channels!!

# Just CD3+ counts: (i.e. first remove dead cells, then calculate CD3+ proportion of live cells.)
```{r}
only.live <- fsApply(trans.fs, function(x) x[which(exprs(x)[, "V450-A"] < 1.5)])
cd3counts <- fsApply(only.live, function(x) length(which(exprs(x)[, "R780-A"] > 1)))/fsApply(only.live, nrow)
t.test(cd3counts[which(full.survival.data > 1000)], cd3counts[which(full.survival.data < 1000)])
boxplot(cd3counts[which(full.survival.data > 1000)], cd3counts[which(full.survival.data < 1000)], boxwex=0.2)
```

